BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-01-25 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
use rmarkdown not latex to insert imgs

    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org]

@@

\includegraphics[height=\textheight, angle=270]{../graphics/plots/workflow.png}

@@

We anticipate that for the typical user BarnebyLives will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 90m resolution is ~16 GiB), while the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of BarnebyLives, generating the labels, requires working installations of Rmarkdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal. 
Unfortunately we have not found Windows alternatives to pdfam and pdftek so we are unable to offer the final label generating functionality on that operating system. 
BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botanists. 
Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry.

## Usage
All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over *3*00 real-world examples are on a Google Sheets accessible from the page.
BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
Virtually all the on-disk memory associated with these data are used in storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation. 
Functions which require the on-disk data require a path to the data as an argument. 
Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.

We anticipate BarnebyLives usage will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 90m resolution is ~16 GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal.
Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.

@@

\begin{wrapfigure}{R}{0.35\textwidth}  
 \begin{center}
    \includegraphics[width=.35\textwidth]{../graphics/plots/collections_map-trim.png}  
  \caption{The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.} 
\end{center}
\end{wrapfigure}

@@

*The number of author abbreviations which were not in the appropriate format were XX (% percent), in nearly all cases the presence or absence of a period were the issue.*
`r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).

@@

\begin{wrapfigure}{R}{0.45\textwidth}
  \begin{center}
    \includegraphics[width=0.40\textwidth]{../graphics/tables/Screenshot-DataSources.png}
  \caption{Sources of Data required for operations}
  \end{center}
\end{wrapfigure}

>>>>

    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$

@@

![Recommended workflow](../graphics/plots/workflow.png){width=100%}

@@

We anticipate that for the typical user BarnebyLives will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 90m resolution is ~16 GiB), while the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of BarnebyLives, generating the labels, requires working installations of Rmarkdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal. 
Unfortunately we have not found Windows alternatives to pdfam and pdftek so we are unable to offer the final label generating functionality on that operating system. 
BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botanists. 
Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry. 
Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.

@@

![The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.](../graphics/plots/collections_map-trim.png){width=50%}

@@

`r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).

@@

![Data Sources](../graphics/tables/Screenshot-DataSources.png){width=50%}
